---
title: "Expanding the xpass Model with Play-Calling Sequences and Tracking Data"
output: html_notebook
---

# Purpose

To expand the pre-existing model for Expected Dropback (`xPass`) to better
detect patterns in offensive play-calling by adding in additional features
at varying levels-of-detail (LODs):

-   the static context of the individual play: personnel, formation, etc.
-   the cumulative history up to this point in the game (and possibly
    season): pass-rush splits, success rates, etc.
-   the pre-snap, frame-by-frame coordinates of players from the tracking
    data

# Setup

Load in datasets

```{r}
source("./r/01_load_data.R")
```

## Libraries

```{r message=FALSE, warning=FALSE}
library(tidyverse)
```

## Data

### Raw and Pre-processed Input

```{r message=FALSE, warning=FALSE}
# non-tracking data
games <- readr::read_csv('./input/games.csv')
players <- readr::read_csv('./input/players.csv')
plays <- readr::read_csv('./input/plays.csv')
player_play <- readr::read_csv('./input/player_play.csv')

# nflverse data
nflverse_joined <- readr::read_tsv("./input/nflverse_joined.tsv")

# modeling dataset
model_data <- readr::read_csv('./input/model_data.csv')
```

# Model

Load in the pre-processed dataset

```{r}
model_data
```

Save a copy of the non-feature columns (week, gameId, playId, frameId,
possessionTeam, label) for joining back in later

```{r}
non_feature_data <- 
   model_data |> 
   select(week, gameId, playId, frameId, possessionTeam, label) |> 
   mutate(rowId = row_number())
```

Minor data cleaning before passing into a model

```{r}
# only keep the presnap versions of the cumulative count columns
model_data_01 <-
   model_data |> 
   mutate(n_plays = pre_pass + pre_rush) |> 
   mutate(presnap_pass_rt = pre_pass / (pre_rush + pre_pass),
          presnap_success_rt = pre_success / n_plays) |> 
   select(-success, -pass_success, -rush_success, 
          -contains("csum"),
          -offenseFormation,
          -pre_pass,
          -pre_success,
          -pre_rush,
          -pre_pass_success,
          -pre_rush_success,
          -pre_success_prop_rush) |> 
   select(-matches(names(non_feature_data)))
```

```{r}
# distinct version to prevent overfitting from using same play multiple times
model_data_01_d <-
   non_feature_data |> 
   bind_cols(model_data_01) |> 
   mutate(
      across(
         .cols = c(RB_min_x, QB_min_x, TE_min_abs_y),
         .fns = function(x) round(x, digits = 0))) |> 
   select(-frameId, -rowId) |> 
   distinct()
```

```{r}
non_feature_data_d <-
   model_data_01_d |> 
   select(week, gameId, playId, possessionTeam, label) |> 
   mutate(rowId = row_number())
```

```{r}
model_data_y_labels_d <- non_feature_data_d$label 

train_row_ids_d <- which(non_feature_data_d$week <= 8)
test_row_ids_d <- which(non_feature_data_d$week == 9)
```

```{r}
model_data_scaled_d <- 
   scale(
      as.matrix(
         model_data_01_d |> 
            select(-matches(names(non_feature_data_d)))))
```

```{r}
X_train_d <- model_data_scaled_d[train_row_ids_d, ]
X_test_d <- model_data_scaled_d[test_row_ids_d, ]

y_train_d <- model_data_y_labels_d[train_row_ids_d]
y_test_d <- model_data_y_labels_d[test_row_ids_d]

scale_pos_weight_d <- sum(y_train_d == 0) / sum(y_train_d == 1)
```

```{r}
# Initialize and Train XGBoost Model
xgb_params_d <- list(
   objective = "binary:logistic",
   scale_pos_weight = scale_pos_weight_d,
   reg_lambda = 300
) 
```

```{r message=FALSE, warning=FALSE}
model_d <- xgboost::xgboost(data = X_train_d,
                            label = y_train_d,
                            params = xgb_params_d,
                            nrounds = 100)
```

```{r}
# Feature Importance
importance_d <- xgboost::xgb.importance(model = model_d) 

feature_importance_df_d <- 
   as_tibble(importance_d) |> 
   arrange(desc(Gain)) 
```

```{r}
# Model Evaluation
y_pred_d <- predict(model_d, X_test_d) |> round() 

conf_matrix_d <- caret::confusionMatrix(factor(y_pred_d), factor(y_test_d))
```

```{r}
# K-Fold Validation
set.seed(904) 

folds_d <- caret::createFolds(model_data_y_labels_d, k = 10) 

cv_scores_d <- purrr::map_dbl(folds_d, function(test_indices) {
   
   train_indices <- setdiff(seq_along(model_data_y_labels_d), test_indices)
   
   model_cv <- xgboost::xgboost(data = model_data_scaled_d[train_indices,],
                                label = model_data_y_labels_d[train_indices],
                                params = xgb_params_d,
                                nrounds = 100,
                                verbose = 0)
   
   pred <- predict(model_cv, model_data_scaled_d[test_indices,]) |> round()
   
   mean(pred == model_data_y_labels_d[test_indices])
   
})
```

```{r}
y_train_pred_d <- predict(model_d, X_train_d) 

y_test_pred_d <- predict(model_d, X_test_d) 
```

```{r}
roc_train_d <- pROC::roc(y_train_d, y_train_pred_d) 
roc_test_d <- pROC::roc(y_test_d, y_test_pred_d) 
```

## Model Comparison

Create predictions on entire dataset

```{r}
model_predictions <- predict(model_d, model_data_scaled_d)
```

Compare to original xPass model

```{r}
xPass_comparison_df <- 
   tibble(gameId = model_data_01_d$gameId,
          playId = model_data_01_d$playId,
          pPass = model_predictions,
          predicted = round(model_predictions),
          actual = model_data_y_labels_d
   ) %>%
   inner_join(nflverse_joined |> 
                 select(gameId, 
                        playId, 
                        team = posteam, 
                        xPass = xpass,
                        pass_OE = pass_oe)) |> 
   mutate(predictedOG = round(xPass)) |> 
   mutate(
      modelCorrect = ifelse(actual == predicted, 1, 0),
      modelCorrectOG = ifelse(actual == predictedOG, 1, 0),
      predictionDiff = pPass - xPass
   ) |> 
   relocate(gameId, playId, team, pPass, xPass, predictionDiff, 
            predicted, predictedOG, actual,
            modelCorrect, modelCorrectOG)

xPass_comparison_df <- 
   xPass_comparison_df |> 
   mutate(bestModel = 
             case_when(
                modelCorrect > modelCorrectOG ~ "new", 
                modelCorrect < modelCorrectOG ~ "OG",
                modelCorrect == modelCorrectOG & 
                   actual == 1 &
                   pPass > xPass ~ "new",
                modelCorrect == modelCorrectOG & 
                   actual == 1 &
                   pPass < xPass ~ "OG",
                modelCorrect == modelCorrectOG & 
                   actual == 0 &
                   pPass < xPass ~ "new",
                modelCorrect == modelCorrectOG & 
                   actual == 0 &
                   pPass > xPass ~ "OG",
                TRUE ~ "other")) |> 
   mutate(bestModelAmount = abs(predictionDiff)) |> 
   mutate(across(.cols = c(predicted, predictedOG, actual),
                 .fns = function(x) ifelse(x == 0, "rush", "pass"),
                 .names = "{.col}_fct")) |> 
   inner_join(nflverse_joined |> 
                 select(gameId, playId, defteam, 
                        score_differential, down, 
                        qtr, ydstogo, epa, wp, wpa, 
                        half_seconds_remaining, 
                        distanceToGoal = yardline_100)) |> 
   inner_join(plays |> 
                 select(gameId, playId, offenseFormation))
```

```{r}
# dataset with model predictions for all frames
preds_all_frames <- 
   model_data_01 |> 
    bind_cols(non_feature_data) |> 
    mutate(across(.cols = c(RB_min_x, QB_min_x, TE_min_abs_y),
                  .fns = function(x) round(x, digits = 0))) |> 
    left_join(model_data_01_d |> mutate(pPass = model_predictions),
              by = join_by(xpass, 
                           OL_count, 
                           QB_count, 
                           RB_count, 
                           TE_count, 
                           WR_count, 
                           RB_min_x, QB_min_x, TE_min_abs_y, 
                           n_plays, 
                           pre_pass_success_rt, 
                           pre_success_prop_pass, 
                           pre_rush_success_rt, 
                           presnap_pass_rt, 
                           presnap_success_rt, 
                           week, 
                           gameId, 
                           playId, 
                           possessionTeam,
                           label))
```

```{r}
# save result
readr::write_csv(preds_all_frames, "./input/predictions_all_frames.csv")
```

Join results together, only keep final prediction

```{r}
preds_final_frames_joined <- 
   preds_all_frames |> 
   rename(team = possessionTeam,
          xPass = xpass) |> 
   mutate(xPass = round(xPass, digits = 4)) |> 
   select(-frameId, -rowId) |> 
   distinct() |> 
   group_by(gameId, playId, team, label) |> 
   slice_tail(n = 1) |> 
   ungroup() |> 
   inner_join(
      xPass_comparison_df |> 
         rename(label = actual) |> 
         mutate(xPass = round(xPass, digits = 4)) |> 
         group_by(gameId, playId, team, label) |> 
         slice_tail(n = 1) |> 
         ungroup()
      ) |> 
   relocate(week, gameId, playId, team, label, pPass, xPass) |> 
   arrange(week, gameId, playId)
```

```{r}
preds_final_frames_joined
```

# Deprecated

## Original Model (deprecated)

```{r}
model_data_y_labels <- model_data$label

train_row_ids <- which(model_data$week <= 8)
test_row_ids <- which(model_data$week == 9)
```

```{r}
X_train <- model_data_scaled[train_row_ids, ]
X_test <- model_data_scaled[test_row_ids, ]

y_train <- model_data_y_labels[train_row_ids]
y_test <- model_data_y_labels[test_row_ids]

scale_pos_weight <- sum(y_train == 0) / sum(y_train == 1)
```

Scale the predictors and then store as matrix in order to use in xgboost

```{r}
model_data_scaled <- scale(as.matrix(model_data_01))
#colnames(model_data_scaled) <- colnames(model_data_01)

```

```{r}
X_train <- model_data_scaled[train_row_ids, ]
X_test <- model_data_scaled[test_row_ids, ]

y_train <- model_data_y_labels[train_row_ids]
y_test <- model_data_y_labels[test_row_ids]

scale_pos_weight <- sum(y_train == 0) / sum(y_train == 1)
```

```{r}
# Initialize and Train XGBoost Model
xgb_params <- list(
   objective = "binary:logistic",
   scale_pos_weight = scale_pos_weight,
   reg_lambda = 300
)
```

```{r}

model <- xgboost(data = X_train,
                 label = y_train,
                 params = xgb_params,
                 nrounds = 100)
```

### Evaluate

```{r}
# Feature Importance
importance <- xgb.importance(model = model)

feature_importance_df <- 
   as_tibble(importance) %>%
   arrange(desc(Gain))
```

```{r}
feature_importance_df
```

```{r}
# Model Evaluation
y_pred <- predict(model, X_test) %>% round()

conf_matrix <- caret::confusionMatrix(factor(y_pred), factor(y_test))
```

```{r}
# K-Fold Validation
set.seed(904)

folds <- caret::createFolds(model_data_y_labels, k = 10)

cv_scores <- map_dbl(folds, function(test_indices) {
   
   train_indices <- setdiff(seq_along(model_data_y_labels), test_indices)
   
   model_cv <- xgboost(data = model_data_scaled[train_indices,],
                       label = model_data_y_labels[train_indices],
                       params = xgb_params,
                       nrounds = 100,
                       verbose = 0)
   
   pred <- predict(model_cv, model_data_scaled[test_indices,]) %>% round()
   
   mean(pred == model_data_y_labels[test_indices])
   
})

```

```{r}
y_train_pred <- predict(model, X_train)

y_test_pred <- predict(model, X_test)
```

```{r}
roc_train <- pROC::roc(y_train, y_train_pred)
roc_test <- pROC::roc(y_test, y_test_pred)
```

## Functions (deprecated)

```{r}
# extracts position numbers from nflverse personnel columns
extract_personnel_count <- function(string, pos_name, default_NA = 0L) {

   # regex pattern to match a digit preceding the string pos_name
   p <- paste0("\\d[ ]", pos_name)


   # extracts the digit as a numeric value
   res <-
      as.numeric(
         stringr::str_extract(
            string = stringr::str_extract(string = string, pattern = p),
            pattern = "\\d")
      )


   # replace NA (no matching pos_name) with default
   # ret <- ifelse(is.na(res), default_NA, res)
   ret <- tidyr::replace_na(res, default_NA)


   return(ret)

}
```

```{r}
# basic euclidean distance
calculate_distance <- 
   
   function(x1, y1, x2, y2) { 
      
      dx = x1 - x2
      
      dy = y1 - y2
      
      sqrt(dx^2 + dy^2)
      
   }
```

```{r}
find_kth_closest <- function(data, point_x, point_y, k) {
   
  # Input validation
  if (!all(c("nflId", "x", "y") %in% colnames(data))) {
    stop("Data frame must contain columns: 'nflId', 'x', and 'y'")
  }
  
  if (!is.numeric(point_x) || !is.numeric(point_y)) {
    stop("point_x and point_y must be numeric values")
  }
  
  if (!is.numeric(k) || k < 1 || k > nrow(data)) {
    stop("k must be a positive integer not exceeding the number of points in the dataset")
  }
   
  
  # Calculate distances from the given point to all points in the dataset
  result <- 
     data |> 
     mutate(
        distance = calculate_distance(point_x, point_y, x, y)
        ) |> 
     arrange(distance) |> 
     slice(k)  # Select the k'th closest point
    
  return(result)
  
}
```

```{r}
# trim excess frames from tracking data
trim_tracking_frames <- 
   
   function(df_tracking, 
            start_events = c("line_set", "man_in_motion"), 
            end_events = c("ball_snap",
                           "autoevent_ballsnap",
                           "snap_direct"),
            keep_cols = TRUE) {
      
      start_frames <- 
         df_tracking |> 
         distinct(gameId, playId, frameId, event) |> 
         group_by(gameId, playId) |> 
         filter(sum(event %in% start_events, na.rm = T) > 0) |> 
         filter(sum(event %in% end_events, na.rm = T) > 0) |> 
         mutate(frameIdStart = min(frameId[event %in% start_events], na.rm = T),
                frameIdEnd = max(frameId[event %in% end_events], na.rm = T)) |> 
         ungroup() |> 
         distinct(gameId, playId, frameIdStart, frameIdEnd)
      
      
      trimmed_tracking <- 
         df_tracking |> 
         inner_join(start_frames, by = join_by(gameId, playId)) |> 
         filter(frameId >= frameIdStart,
                frameId <= frameIdEnd) 
      
      if (keep_cols == FALSE) {
         
         trimmed_tracking$frameIdStart <- NULL
         trimmed_tracking$frameIdEnd <- NULL
         
      } 
      
      return(trimmed_tracking)
      
   }
```

```{r}
# determine who the center is on a play
determine_center_player2 <- 
   
   function(df_tracking, 
            df_plays = plays, 
            df_players = players, 
            snapFrameOffset = NULL) {
   
   # player positions to exclude from being classified as the center
   ineligible_players <- 
      df_players |> 
      filter(position %in% c("QB","WR","RB","FB","TE")) |> 
      distinct(nflId, displayName, position)
   
   
   coords_at_snap <- 
      df_tracking |> 
      trim_tracking_frames() |> 
      distinct(gameId, playId, club, nflId, displayName, frameId, event, x, y)
   
   
   fb_xy <- 
      coords_at_snap |> 
      filter(club == "football") |> 
      select(gameId, playId, frameId, x_fb = x, y_fb = y)
   
   
   team_sides <- 
      df_plays |> 
      distinct(gameId, playId, possessionTeam, defensiveTeam)
   
   
   coords_at_snap <- 
      coords_at_snap |> 
      filter(club != "football") |> 
      left_join(fb_xy, 
                 by = join_by(gameId, playId, frameId)) |> 
      left_join(team_sides, 
                 by = join_by(gameId, playId)) |> 
      mutate(clubSide = case_when(club == possessionTeam ~ "OFF",
                                  club == defensiveTeam ~ "DEF",
                                  club == "football" ~ "FB",
                                  TRUE ~ "ERR"))
   
   
   coords_at_snap |> 
      mutate(dist_to_fb = calculate_distance(x, y, x_fb, y_fb)) |> 
      filter(clubSide == "OFF") |> 
      anti_join(ineligible_players, by = join_by(nflId)) |> 
      arrange(gameId, playId, dist_to_fb) |> 
      group_by(gameId, playId, frameId) |> 
      slice_head(n = 1) |> 
      ungroup()
   
}
```

```{r}
determine_linemen <- 
   
   function(df_tracking, 
            df_plays = plays, 
            df_players = players) {
   
   tracking_center <- 
      df_tracking |> 
      determine_center_player2(df_plays = plays, df_players = players) |> 
      select(gameId, 
             playId, 
             frameId, 
             event, 
             club, 
             nflId_C = nflId, 
             displayName_C = displayName, 
             x_C = x, 
             y_C = y) #|> mutate(position = "C")
   
   
   df_players <- df_players |> distinct(nflId, displayName, position)
   
   # player positions to exclude from being classified as linemen
   ineligible_players <- 
      df_players |> 
      filter(position %in% c("QB","WR","RB","FB","TE")) |> 
      distinct(nflId)
   
   # tracking of "eligible" players only aka OL
   tracking_OL <- 
      df_tracking |> 
      inner_join(tracking_center) |> 
      filter(nflId_C != nflId) |> 
      #inner_join(df_tracking, by = join_by(gameId, playId, frameId, event, club)) |> 
      anti_join(ineligible_players, by = join_by(nflId)) |> 
      select(gameId, 
             playId, 
             frameId, 
             event, 
             club, 
             nflId, 
             displayName, 
             x,
             y,
             x_C,
             y_C) |> 
      arrange(gameId, 
              playId, 
              club,
              frameId, 
              nflId)
   
   # add helpers to determine on-field position
   res <- 
      tracking_OL |> 
      mutate(sideOfC = ifelse(y > y_C, "L", "R")) |> 
      mutate(distToC = calculate_distance(x, y, x_C, y_C)) |> 
      arrange(gameId, playId, frameId, sideOfC, distToC) |> 
      group_by(gameId, playId, frameId, club, sideOfC) |> 
      mutate(distRank = seq_along(nflId)) |> 
      ungroup()
   
   
   return(res)
   
   
}
```

Test metric - squared horizontal distance from the LT and RTs. If summed up
all players values on a play, then higher values should imply higher xPass?

```{r}
widths_from_tackles <- 
   
   function(df_tracking, 
            df_plays = plays, 
            df_players = players,
            buffer_distance = 0) {
      
      y_bounds <- 
         df_tracking |> 
         determine_linemen() |> 
         filter(distRank == 2) |> 
         group_by(gameId,playId,frameId,event,club) |> 
         summarise(y_RT = min(y, na.rm = T),
                   y_LT = max(y, na.rm = T)) |> 
         ungroup() |> 
         mutate(y_RTo = y_RT - buffer_distance,
                y_LTo = y_LT + buffer_distance)
      
      tracking_outside_of_y <- 
         df_tracking |> 
         inner_join(y_bounds, by = join_by(gameId,playId,frameId,event,club)) |> 
         mutate(isOutsideTackles = case_when(y < y_RTo ~ "OUTSIDE",
                                             y <= y_LTo ~ "INSIDE",
                                             y > y_LTo ~ "OUTSIDE",
                                             TRUE ~ "ERR")) |> 
         mutate(dy2 = case_when(isOutsideTackles == "INSIDE" ~ 0,
                                y < y_RTo ~ (y_RT - y)^2,
                                y > y_LTo ~ (y - y_LT)^2,
                                TRUE ~ NA_real_))
         #filter(isOutsideTackles == "OUTSIDE")
         #filter(y < y_RT | y > y_LT)
      
      return(tracking_outside_of_y)
      
   }
```

```{r}
determine_position_from_alignment <- 
   
   function(df_tracking, df_plays = plays, df_players = players) {
      
      
      center_info <- 
         df_tracking |> 
         determine_center_player2(df_plays = df_plays, df_players = df_players) |> 
         select(gameId, playId, club, frameId, x_c = x, y_c = y)
      
      
      joined_tracking <- 
         df_tracking |> 
         inner_join(center_info, by = join_by(gameId, playId, frameId, club)) |> 
         relocate(gameId, 
                  playId, 
                  club, 
                  nflId, 
                  displayName, 
                  frameId, 
                  x, 
                  y, 
                  x_c, 
                  y_c)
      
      
      joined_tracking$dx <- joined_tracking$x - joined_tracking$x_c
      joined_tracking$dy <- joined_tracking$y - joined_tracking$y_c
      
      
      joined_tracking |> 
         
         mutate(sideOfCenter = 
                   case_when(abs(dy) <= 1 ~ "C",
                             dy > 0 ~ "L", 
                             dy < 0 ~ "R",
                             TRUE ~ NA_character_)) |> 
         
         mutate(widthLevel = 
                   case_when(
                      sideOfCenter == "C" ~ "C",
                      abs(dy) <= 4.5 ~ "BOX",
                      abs(dy) > 4.5 ~ "WIDE",
                      TRUE ~ "OTHER"
                   )
         ) |> 
         
         mutate(depthFromCenter = 
                   case_when(
                      # begin with labeling the center, QB, and RBs (if any)
                      dx == 0 & dy == 0 ~ "C",
                      abs(dy) <= 1 & abs(dx) <= 1.5 ~ "UNDER_CENTER",
                      abs(dy) <= 1 & abs(dx) > 1.5 ~ "BACKFIELD",
                      # then label players in the box (OL)
                      abs(dy) <= 3.5 & abs(dx) <= 1.5 ~ "OL",
                      # then eligible receivers
                      abs(dy) > 3.5 ~ "REC",
                      TRUE ~ "OTHER"
                   )
         ) |> 
         
         arrange(gameId, 
                 playId, 
                 frameId, 
                 sideOfCenter, 
                 depthFromCenter, 
                 desc(abs(dy))) |> 
         
         group_by(gameId, 
                  playId, 
                  frameId,
                  sideOfCenter, 
                  depthFromCenter) |> 
         
         mutate(alignmentNum = ifelse(sideOfCenter == "C", 
                                      0L, 
                                      seq_along(nflId))) |> 
         ungroup()
   

}
```

Example usage:

```{r}
tracking |> 
    filter(gameId == 2022091200, playId == 315) |> 
    determine_position_from_alignment() |> 
    arrange(gameId, playId, club, nflId, frameId) |> 
    filter(frameId %in% seq(170, 200, 5)) |> 
    group_by(gameId, playId, frameId) |> 
   mutate(
      n_backfield = sum(depthFromCenter == "BACKFIELD", na.rm = TRUE),
      n_wide_eligible = 
         sum(depthFromCenter == "REC" & widthLevel == "WIDE", na.rm = TRUE),
      n_box_eligible = 
         sum(depthFromCenter == "REC" & widthLevel == "BOX", na.rm = TRUE)) |> 
    add_count(depthFromCenter, name = "n_depth") |> 
    add_count(widthLevel, name = "n_width") |> 
    ungroup() |> 
    arrange(gameId, playId, club, frameId, depthFromCenter, widthLevel, dy) |> 
   relocate(gameId, playId, frameId, nflId, displayName, jerseyNumber,
            x, y, dx, dy, 
            sideOfCenter:alignmentNum,
            starts_with("n_"))
```
